%version of 07-22-20

\chapter*{Manifesto}

The technologies that enable both the hardware and software systems of modern computers have grown in complexity at least as fast as they have in performance.  For decades, it has been impossible to design computing systems without using tools whose underpinnings are rooted
in {\it discrete mathematics}.  For too long, though, these tools have been used to create widgets which are then assembled into coherent systems using only cleverness and untutored ingenuity.  The net result is that many of the computing systems which impact our daily lives are neither well-understood nor well-controlled.  We read daily about breaches of security and/or privacy that can neither be controlled nor repaired because the victimized system is too complex to be understood.  The lesson is clear: {\bf We must stop being satisfied with mere {\em knowledge} of how to make a system take in inputs and emit outputs.  We must strive for {\em understanding} of every system's structure and behavior.}  There are myriad examples of successful such computing systems: compilers and databases and electronic circuits begin a long list.  Notably, many success stories trace back to the development of relevant new discrete mathematics.

This book has a simple but fundamental goal---to endow readers with {\em operational} conceptual and methodological understanding of the discrete mathematics that can be used to study, and understand, and perform computing.  We want each reader to {\em understand} the elements of computing, rather than just {\em know} them.  Thereby, the interested reader will be able to {\em develop concepts} and {\em invent techniques and technologies} for expanding the capabilities of the hardware that performs computations and the software that controls the hardware.  We stress the word {\it operational:}  {\bf We want the readers' level of understanding to enable them to {\em ``do''} mathematics.}

\bigskip

\index{Kronecker, Leopold}

The bold goal of reaching the point of ``doing'' mathematics is within the reader's grasp.  We brace the spine and embolden the spirit by invoking no less an authority than the great mathematician Leopold Kronecker.  We read in the historical tome \cite{Bell86} of Kronecker's assurance that ``God made the integers; all else is the work of man.''  Therefore, in order to achieve our goal of ``doing'' mathematics, we need only stand on the shoulders of (mathematical) giants---there is no need to insolently attempt to wrest fire from Olympus!

\ignore{***************
\bigskip

\noindent \fbox{
\begin{minipage}{0.95\textwidth}
The phrase ``standing on the shoulders of giants'' has been cited by
many, over many centuries, ranging from Sir Isaac Newton to Bernard of
Chartres, and beyond.  The treatise by Merton \cite{Merton}
\index{Merton, Robert K.} provides an entertaining, educational
history of the phrase.
\end{minipage}
}

****************}


