%version of 01-14-20
\documentclass{article}[12pt]

\begin{document}

\begin{center}
BOOK PROPOSAL 

\bigskip

{\large\bf\em UNDERSTAND MATHEMATICS,

UNDERSTAND COMPUTING}
\end{center}

\begin{center}
\begin{tabular}{ccc}
{\em Arnold L.~Rosenberg} & & {\em Denis Trystram} \\
Distinguished Univ.~Professor Emeritus
  & & Distinguished Professor \\
University of Massachusetts  & & University of Grenoble Alpes \\
Amherst, MA 01003, USA       & & Grenoble, FRANCE \\
{\small\tt rsnbrg@cs.umass.edu} & & {\small\tt denis.trystram@univ-grenoble-alpes.fr}
\end{tabular}
\end{center}

\section*{Overview of the Book}

The authors propose a book on discrete mathematics that is novel in many aspects of motivation, organization, and content.
\begin{itemize}
\item
{\em The book is designed to fill multiple needs for both students and instructors}:
  \begin{itemize}
  \item
as a \underline{textbook} for a beginning university discrete math course

  \item
as a \underline{source of mathematical preliminaries} for a broad range of beginning and intermediary university courses:

\smallskip

including algorithms; artificial intelligence; computer architecture and digital design; information processing and retrieval; networking and security; social media; web-based computing

  \item
as a \underline{tutorial} on how mathematicians model and think about the worlds of mathematics, of science, of nature, and of people

  \item
as a \underline{launchpad} into further study of the historical, social, and technical aspects of mathematics
  \end{itemize}
\end{itemize}
{\sf Our focus on mathematics as a human endeavor separates us from many competitors \\
---especially our repeated reminders that human needs and curiosity underlie the development of new mathematics.}

\medskip

\begin{itemize}
\item
{\em The primary goal of the book is to teach the reader how to \underline{do mathematics}.}

\smallskip

To this end, we provide {\em multiple explanations} of key results and concepts, employing {\em multiple representations} for key notions (such as number), and {\em multiple, quite-distinct, proofs} of basic results.

\smallskip

We hope thereby to empower and encourage readers to develop {\em their individual insights and talents} as they look at and reason about the world.  We know that no single presentation to be ``the right one'" for every reader.
\end{itemize}
{\sf This is another important way in which we differ from competing texts.}

\medskip

\begin{itemize}
\item
{\em The book focuses squarely on discrete mathematics as a topic of study.}

\smallskip

However, the authors have called upon their decades-long experience as researchers in discrete mathematics, computer science, and computer engineering to inform their {\em choice of topics and examples} and the {\em organization of the material}.

\smallskip

That said, while the book can serve as {\em preparation} for studying a range of computational topics, it is {\em not an introduction} to such material: The authors have diligently selected mathematical material whose application goes beyond any single focus, even one as broad as computation.
\end{itemize}

\bigskip

\noindent
Our commitment to teaching how to \underline{do mathematics} is discernible in certain unique features in the text.  To cite just three relevant features:
  \begin{itemize}
  \item
The first technical (i.e., non-introductory) chapter is entitled {\em Doing Mathematics}.  This chapter describes and amply illustrates topics such as: {\em the nature of mathematical argumentation}, including techniques of proof; {\em the centrality of representation} for the successful manipulation and analysis of mathematical concepts.

\smallskip

By concentrating this material in a single place---in contrast to introducing it piecemeal, as several competitors do---we help readers recognize that many results can be proved in many ways.  In fact we stress this by proving exemplary results in multiple ways.

 \item
The preceding ``many roads to Rome" philosophy is a hallmark of our text, which separates it from many competitors.  As a ``maximal' example, we evaluate one particular summation in five completely distinct ways. 

%We reinforce this philosophy repeatedly by historical anecdote  The history of mathematics---and even the  authors' personal histories of mathematical research---is full of examples of significant theorems and calculations that long evaded successful solution until investigators discovered the ``right" way to think about a particular aspect of the target problem.

  \item
The book develops the all-important concept of ``number" via a sequence of {\em seven} intellectually interwoven chapters:
   \begin{enumerate}
   \item       
The first chapter describes {\em the basics of our number system}, including the historical needs that led to each successive augmentation of the system.  The differences in properties among the various components of our system (integers vs.~rationals vs.~reals vs.~complexes) are integrated with discussions of topics such as order and divisibility.
  \item
The second chapter ``puts numbers to work" by means of {\em arithmetic operations and important classes of numerical functions}.
  \item
The third chapter develops notions relating to {\em summation}, including techniques for exactly and approximately summing important classes of finite and infinite series.
  \item
The fourth chapter expands on a very important topic that rears its head when one works with infinite mathematical objects---{\em how to deal with infinity}, e.g., with topics such as the ``point at infinity'' and with infinite series.  In our organization of mathematics, the subject of infinite summations is an opportune trigger for discussing infinity.
     \begin{itemize}
     \item
We need infinite geometric sums to talk about positional number systems and their numerals;
     \item
Infinite summations point out dramatically that there are situations in which infinity does not change the way we do mathematics---some infinite summations ``converge", hence behave like numbers---and other situations in which infinity makes a big change---some infinite summations ``diverge".
     \item
Divergent series provide a great opportunity to introduce the crucial topic of {\em asymptotics}, which stays with us during much of the rest of the book.
     \end{itemize}
The chapter closes with a discussion of the various genres of paradox that arise when one works with the infinite.  These illustrate the impact of infinity in non-numerical domains.
  \item
The fifth chapter resumes the study of numbers.  Building on its predecessors, this chapter exposes the intricacies of: the {\em prime numbers} (the building blocks of the integers); {\em pairing functions} (a building block of the encodings that provide much conceptual power as one computes); {\em finite number systems}.
  \item
The sixth chapter is devoted to {\em recurrences}.  Coverage ranges from {\em techniques for solving recurrences}, to {\em important recursively defined sequences} (mainly, the binomial coefficients and Fibonacci numbers), to {\em sophisticated examples and applications}.
  \item
The final chapter in the sequence deals with {\em numerals: representations of numbers}.  Topics include: {\em Classes of ``positional" numerals}---the familiar $b$-ary numerals, the less-familiar $b$-adic numerals, the ``carry-free" signed-digit numerals.  The chapter also covers {\em proofs of uncountability} (which brings infinity back to our attention).  
  \end{enumerate}

\item
Our long sequence of chapters on numbers and number systems illustrates yet another philosophical pillar of our approach to teaching/learning mathematics.  We have structured the coverage of important topics---numbers, sets, graphs, combinatorics, etc.---so that a reader can choose a level of coverage that will make any specific topic a ``casual acquaintance" or a ``good friend" or an ``intimate friend".

\smallskip

And, the reader can return at a later date---days or months later, or even years later---and intensify the relationship!
\end{itemize}

\begin{itemize}
\item
Every topic is treated as a new vehicle for mathematical discovery and reasoning.  The recurring subject of binomial coefficients is a good example.  This class of integers is an important exemplar of {\em the power of recursive definitions}.  By their appearance in the Binomial Theorem, these numbers illustrate {\em the power of compact representations}---for both reasoning and computation.  By dint of their multiple uses within the study of combinatorics and probability, the numbers {\em demonstrate completely unexpected interrelationships among seemingly unrelated concepts}.

\item
Mathematics is invented by humans.  Motivated by superficial curiosity, humans exploit their innate powers to detect patterns.  Motivated by a deeper form of curiosity, humans develop conceptual tools in order to explicate the observed patterns.

\smallskip

We continually strive to lend the reader a historical perspective on the way that progress has been made in major areas in mathematics.  By exposing the human side of what have turned out to be fruitful pursuits of curiosity, we strive to motivate readers to launch their own journeys of discovery.  Illustrative examples of fruitful curiosity:

Archimedes wondered how many grains of sand filled beaches; Leonardo Pisano (Fibonacci) was curious about the demographics of rabbits; John Arbuthnot wondered whether male births really outnumbered female births among humans. 
\end{itemize}

\section*{The Authors}

\subsection*{Research}

The authors have a long cumulative history of research on
\begin{description}
\item[{\sf Mathematics:}]
discrete math and mathematical logic
\item[{\sf Computer Science:}]
algorithms, computer architecture, computation theory, data analytics, information retrieval, resource management, task scheduling
\item[{\sf Computer Engineering:}]
digital logic design, energy conservation, fault tolerance, testing of digital logic
\end{description}

\smallskip

\noindent
They have cumulatively published more than 450 research articles in elite journals and conferences, as well as several research books and a few textbooks on both mathematical and computational subjects.  %(Our attached CV's supply detailed numbers and titles.)

% I have about 100 journals + 130 conferences + 12 books, according to DBLP, you have 120 journals and about 100 conferences
\smallskip

\subsection*{Teaching}

Over a cumulative span approaching 3/4 century, the authors have taught courses ranging from mathematics and mathematical logic to a broad spectrum of topics in computer science and engineering at the following universities (listed alphabetically):

\smallskip

\noindent
Duke University, 
University of Grenoble,
University of Massachusetts,
University of Paris,
New York University (NYU),
Polytechnic Institute of New York,
The Technion (Israel Institute of Technology),
Tunis Scientific University,
University of Toronto,
Yale University

\medskip

\noindent
The courses taught by the authors include:

\smallskip

\noindent
Algorithms,
Computation, Complexity, and Approximation Theory,
Computer Architecture,
Mathematical Logic,
Research Methodology,
VLSI Design

\end{document}

\section*{The Book: What Is Special}

Focus is not just on {\em learning topics in mathematics} but in learning {\em how to ``do mathematics"} --- how to {\em think like a mathematician} in terms of {\em problem solving} and {\em formulating ``real" problems mathematically}.  Here are some steps we take:
\begin{itemize}
\item
We have a chapter entitled ``Doing Mathematics".  We are not claiming that this is unique.  But this is one of the steps that we take.
\item
We provide numerous rather different explanations---including proofs---when we feel that the reader may benefit from learning multiple ways to think about the same issue.
\item
We strive to transmit important pieces of mathematical culture by telling stories that expose how mathematical greats of the past came to their discoveries.
\item
We go ``beyond the basics" so that the reader who is fascinated by a particular topic can 		test the waters" with more advanced and/or specialized material than one typically encounters in an introductory course.  We even include a small number of essays---in the appendix---that really give a peek into specialized topics.
\item
We have partitioned our references and our index to help the reader find {\em historical} material and references that are more {\em cultural} than technical (Euclid, Al-Khwarizmi).
\end{itemize}


\begin{itemize}
\item
Topics chosen not only from ``standard" syllabi, but also from our personal experience as researchers.
\item
Topics are organized with an eye toward the mathematical tools needed to master them.

Talk about the sequence: \\
NUMBERS1 \\
ARITHMETIC \\
SUMMATION \\
INFINITE \\
NUMBERS2 \\
RECURRENCES \\
NUMBERS3
\end{itemize}

\end{document}